Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-28 paper.qmd
Refine language on Nix and reproducibility tools
Updated language for clarity and precision in the discussion of Nix and its tools, emphasizing reproducibility and user accessibility.

[Nix]{.pkg} is a cross-platform package manager designed for reproducible
software installation and deployment. Unlike traditional package managers,
[Nix]{.pkg} emphasizes immutable infrastructure and a functional programming
paradigm to ensure computational environment consistency. The primary
[Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
packages, including nearly all of CRAN and Bioconductor. According to Repology,
a service tracking package repositories across distributions, `nixpkgs` is both
the largest and among the most up-to-date repositories available (Figure
@fig-repology).

@@

Many tools exist to improve reproducibility, but [Nix]{.pkg} stands out because
it deploys complete software environments *closed under the "depends on"
relation*: it installs not only a package, but all its dependencies and their
dependencies. This makes [Nix]{.pkg} particularly powerful for reproducible
research.

However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
by providing a familiar interface and workflow. By building reproducible
development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
applications, or serving [plumber]{.pkg} APIs.

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging [Nix]{.pkg} as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of polyglot analyses. Most importantly, it provides native
polyglot support without the manual coordination required by containerization
approaches, making complex multi-language workflows accessible to researchers
without systems administration expertise.

Adopting [Nix]{.pkg}-based workflows represents a significant paradigm shift, and
researchers should approach it incrementally. A reasonable pathway begins with
using [rix]{.pkg} to generate reproducible environments for new projects, while
maintaining existing workflows for ongoing work. This allows learning [Nix]{.pkg}
concepts without risking active research. As comfort grows, researchers can
experiment with [rixpress]{.pkg} for smaller analyses before tackling complex
polyglot pipelines. Institutions can support adoption by providing workshops,
maintaining institutional binary caches with private packages, and designating
local experts. The investment pays increasing returns: early projects require
substantial learning, but subsequent projects leverage accumulated knowledge and
reusable environment definitions. For research groups, having one member develop
[Nix]{.pkg} expertise can enable the entire team to benefit from reproducible
environments without everyone climbing the learning curve. Ultimately, the
decision to adopt these tools depends on project timescales, reproducibility
requirements, and institutional context—but the framework exists for those who
need it.

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide
significant steps forward, it is important to acknowledge current limitations.
Debugging complex [rixpress]{.pkg} pipelines remains more difficult than
debugging [targets]{.pkg} workflows, particularly when failures occur deep in
the [Nix]{.pkg} build process. The visualization capabilities, while useful, lack
features like automatic detection of outdated derivations that make tools like
[targets]{.pkg} particularly developer-friendly. Cross-platform reproducibility,
while generally strong, has known limitations on macOS due to system framework
dependencies. These are not insurmountable problems—they represent areas for
continued development and community contribution. Researchers adopting these
tools should do so with realistic expectations: they provide powerful
capabilities for deep reproducibility and polyglot workflows, but require
patience during the learning phase and may occasionally present challenges that
simpler tools avoid.

>>>>

[Nix]{.pkg} is a cross-platform package manager designed for reproducible
software installation and deployment. Unlike traditional package managers,
[Nix]{.pkg} emphasizes immutable infrastructure and a functional programming
paradigm to ensure computational environment consistency. The primary
[Nix]{.pkg} package collection, `nixpkgs`, provides access to over 120,000
packages, including nearly all of CRAN and Bioconductor.^[Only about 80 packages 
currently fail to build on the NixOS build farm, plus roughly 30 packages marked as broken
by the maintainers of the [R]{.proglang} ecosystem. These could, in principle, be fixed, 
but doing so would require significant effort. Marking them as broken signals to users 
that their build failures are known and intentional.] According to Repology,
a service tracking package repositories across distributions, `nixpkgs` is both
the largest and among the most up-to-date repositories available (Figure
@fig-repology).

@@

Many tools exist to improve reproducibility, but [Nix]{.pkg} stands out because 
it deploys complete software environments closed under the “depends on” relation:
it installs not only a package but all of its transitive dependencies. 
This makes [Nix]{.pkg} uniquely suited for reproducible research.

Yet solving such a complex problem makes [Nix]{.pkg} itself complex, as it relies on a 
functional paradigm still likely unfamiliar to most researchers.
[rix]{.pkg} lowers this barrier for [R]{.proglang} users by providing a familiar interface and workflow. 
By declaratively specifying reproducible development shells with [Nix]{.pkg}, 
researchers can accommodate diverse use cases—from running analytical pipelines 
to developing interactive [shiny]{.pkg} applications or serving [plumber]{.pkg} APIs.

Furthermore, [rixpress]{.pkg} extends this declarative model to entire analysis pipelines. 
By leveraging [Nix]{.pkg} as a unified build automation system, [rixpress]{.pkg} enables polyglot 
workflows where each step runs in a hermetically sealed environment. 
This ensures both spatial and temporal reproducibility, efficient caching, and seamless 
orchestration of multi-language analyses. Crucially, it provides native polyglot support 
without the manual orchestration required by container-based approaches, 
making complex workflows accessible without systems administration expertise.

Adopting [Nix]{.pkg}-based workflows represents a conceptual shift from procedural 
scripting to declarative specification. A practical adoption pathway begins with using 
[rix]{.pkg} to generate reproducible environments for new projects while maintaining existing workflows. 
This allows gradual learning of [Nix]{.pkg} concepts without disrupting active research. 
As confidence grows, researchers can extend to [rixpress]{.pkg} for smaller analyses before tackling complex 
pipelines. Institutions can support adoption by offering workshops, maintaining binary caches, 
and designating local experts. While early projects require investment, later projects benefit from
reusable environment definitions and cumulative expertise. Ultimately, adoption depends on 
project scope, reproducibility needs, and institutional context—but the framework now exists 
for those who require robust, long-term reproducibility.

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) represent significant progress,
current limitations remain. Debugging complex [rixpress]{.pkg} pipelines is more challenging than
troubleshooting [targets]{.pkg} workflows, especially when failures occur deep in the [Nix]{.pkg} build process. 
Visualization capabilities, though useful, currently lack features such as automatic detection 
of outdated derivations. Cross-platform reproducibility remains generally strong but is constrained on macOS by 
proprietary framework dependencies. These challenges are active areas for development and community contribution. 
Researchers adopting these tools should do so with realistic expectations: they provide strong guarantees of spatial 
and temporal reproducibility and scalable polyglot workflows, but they require patience during the learning phase 
and may occasionally present challenges that simpler tools avoid.
